### 2012-11-01 doc/prelim/document/chapters/parallel_mc/parallel_mc.tex
finsihed draft except for conclusion
updated presentation
fixed adjoint expectation value

Using these matrices, we can then form the expectation value of the
direct solution. For a given random walk permutation $\nu$ with $k$
events, we define the weight of that permutation to be:
\begin{equation}
  W_{m} = \sum_{m=0}^k w_{i,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}\:,
  \label{eq:direct_permutation_weight}
\end{equation}
such that the weight of each transition event contributes to the
total. The contribution to the solution from a particular random walk
permutation is then:
\begin{equation}
  X_{\nu}(i_0 = i) = \sum_{m=0}^k W_{m} b_{i_m}\:,
  \label{eq:direct_permutation_contribution}
\end{equation}
where $X_{\nu}(i_0 = i)$ signifies that the solution state in which we
are tallying defines state $i$.  We can interpret this precisely as
before in that during the random walk we collect the source in the
states that are visited and apply them to the solution tally. We then
define the probability that a particular random walk permutation of
$k$ events will occur:
\begin{equation}
  P_{\nu} = p_{i,i_1} p_{i_1,i_2} \cdots p_{i_{k-1},i_k}\:.
  \label{eq:direct_permutation_probability}
\end{equation}
Finally, we define the expectation value of $X$ to be the collection
of all random walk permutations and their probabilities:
\begin{equation}
  E\{X(i_0 = i)\} = \sum_{\nu} P_{\nu} X_{\nu}\:,
  \label{eq:direct_expectation_value}
\end{equation}
which, if expanded, directly recovers the exact solution:
\begin{equation}
  \begin{split}
    E\{X(i_0 = i)\}
    &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
    w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{k-1},i_k} b_{i_k}\\ &= x_i\:,
  \end{split}
  \label{eq:direct_expectation_expansion}
\end{equation}
therefore providing an unbiased Monte Carlo estimator.

@@

In terms of particle transport, this adjoint method is equivalent to a
traditional forward method. As a result of using the adjoint system,
we modify our probabilities and weights using the \textit{adjoint
  Neumann-Ulam decomposition} of $\ve{H}$:
\begin{equation}
  \ve{H}^{T} = \ve{P} \circ \ve{W}\:,
  \label{eq:adjoint_neumann_ulam}
\end{equation}
where now we are forming the decomposition with respect to the
transpose of $\ve{H}$\footnote{This is sometimes referred to as the
  adjoint form for real-valued matrices}. We then follow the same
procedure as the direct method for forming the probability and weight
matrices in the decomposition. Using the adjoint form, probabilities
should instead be column-scaled:
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,
  \label{eq:adjoint_probability}
\end{equation}
such that we expect to select a new state $j$ from the current state
in the random walk $j$ by sampling column-wise. Per
Eq~(\ref{eq:adjoint_neumann_ulam}), the transition weight is then
defined as:
\begin{equation}
  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
  \label{eq:adjoint_weight}
\end{equation}
Using the decomposition we can then define an expectation value for
the adjoint method. Given Eq~(\ref{eq:direct_permutation_weight}) as
the weight generated for a particular random walk permutation as in
Eq~(\ref{eq:mc_walk_permutation}) and our result from
Eq~(\ref{eq:inner_product_constraint}) generated by applying the
adjoint constraints, the contribution to the solution in state $i$
from a particular random walk permutation is then:
\begin{equation}
  X_{\nu} = \sum_{m=0}^k W_{m} b_{i_0} \delta_{i,i_m}\:,
  \label{eq:adjoint_permutation_contribution}
\end{equation}
where the Kronecker delta indicates that the tally contributes only in
the current state and $b_{i_0}$ is the sampled source starting
state. Note here that the estimator in
Eq~(\ref{eq:adjoint_permutation_contribution}) does not have a
dependency on the source state as in
Eq~(\ref{eq:adjoint_permutation_contribution}), providing a remedy for
the situation in the direct method where we must start a random walk
in each source state for every permutation such that we may compute a
contribution for that state. In the adjoint method, we instead tally
in all states and those of lesser importance will not be visited as
frequently by the random walk. Finally, the expectation value using
all permutations is:
\begin{equation}
  E\{X\} = \sum_{\nu} P_{\nu} X_{\nu}\:
  \label{eq:adjoint_expectation_value}
\end{equation}
which, if expanded in the same way as the direct method, directly
recovers the exact solution:
\begin{equation}
  \begin{split}
    E\{X_j\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N} b_{i_0} h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}
    \delta_{i_k,j} \\ &= x_{j}\:,
  \end{split}
  \label{eq:adjoint_expectation_expansion}
\end{equation}
therefore also providing an unbiased Monte Carlo estimate of the
solution.

>>>>

Using these matrices, we can then form the expectation value of the
direct solution. For a given random walk permutation $\nu$ with $k$
events, we define the weight of that permutation to be:
\begin{equation}
  W_{m} = w_{i,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}\:,
  \label{eq:direct_permutation_weight}
\end{equation}
such that the weight of each transition event contributes to the
total. The contribution to the solution from a particular random walk
permutation is then:
\begin{equation}
  X_{\nu}(i_0 = i) = \sum_{m=0}^k W_{m} b_{i_m}\:,
  \label{eq:direct_permutation_contribution}
\end{equation}
where $X_{\nu}(i_0 = i)$ signifies that the solution state in which we
are tallying defines state $i$.  We can interpret this precisely as
before in that during the random walk we collect the source in the
states that are visited and apply them to the solution tally. We then
define the probability that a particular random walk permutation of
$k$ events will occur:
\begin{equation}
  P_{\nu} = p_{i,i_1} p_{i_1,i_2} \cdots p_{i_{k-1},i_k}\:.
  \label{eq:direct_permutation_probability}
\end{equation}
Finally, we define the expectation value of $X$ to be the collection
of all random walk permutations and their probabilities:
\begin{equation}
  E\{X(i_0 = i)\} = \sum_{\nu} P_{\nu} X_{\nu}\:,
  \label{eq:direct_expectation_value}
\end{equation}
which, if expanded, directly recovers the exact solution:
\begin{equation}
  \begin{split}
    E\{X(i_0 = i)\}
    &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
    w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{k-1},i_k} b_{i_k}\\ &= x_i\:,
  \end{split}
  \label{eq:direct_expectation_expansion}
\end{equation}
therefore providing an unbiased Monte Carlo estimator.

@@

In terms of particle transport, this adjoint method is equivalent to a
traditional forward method. As a result of using the adjoint system,
we modify our probabilities and weights using the \textit{adjoint
  Neumann-Ulam decomposition} of $\ve{H}$:
\begin{equation}
  \ve{H}^{T} = \ve{P} \circ \ve{W}\:,
  \label{eq:adjoint_neumann_ulam}
\end{equation}
where now we are forming the decomposition with respect to the
transpose of $\ve{H}$\footnote{This is sometimes referred to as the
  adjoint form for real-valued matrices}. We then follow the same
procedure as the direct method for forming the probability and weight
matrices in the decomposition. Using the adjoint form, probabilities
should instead be column-scaled:
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,
  \label{eq:adjoint_probability}
\end{equation}
such that we expect to select a new state $j$ from the current state
in the random walk $j$ by sampling column-wise. Per
Eq~(\ref{eq:adjoint_neumann_ulam}), the transition weight is then
defined as:
\begin{equation}
  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
  \label{eq:adjoint_weight}
\end{equation}
Using the decomposition we can then define an expectation value for
the adjoint method. Given Eq~(\ref{eq:direct_permutation_weight}) as
the weight generated for a particular random walk permutation as in
Eq~(\ref{eq:mc_walk_permutation}) and our result from
Eq~(\ref{eq:inner_product_constraint}) generated by applying the
adjoint constraints, the contribution to the solution in state $i$
from a particular random walk permutation is then:
\begin{equation}
  X_{\nu} = \sum_{m=0}^k W_{m} \delta_{i,i_m}\:,
  \label{eq:adjoint_permutation_contribution}
\end{equation}
where the Kronecker delta indicates that the tally contributes only in
the current state and $b_{i_0}$ will be the sampled source starting
weight. Note here that the estimator in
Eq~(\ref{eq:adjoint_permutation_contribution}) does not have a
dependency on the source state as in
Eq~(\ref{eq:adjoint_permutation_contribution}), providing a remedy for
the situation in the direct method where we must start a random walk
in each source state for every permutation such that we may compute a
contribution for that state. In the adjoint method, we instead tally
in all states and those of lesser importance will not be visited as
frequently by the random walk. Finally, the expectation value using
all permutations is:
\begin{equation}
  E\{X\} = \sum_{\nu} P_{\nu} X_{\nu}\:
  \label{eq:adjoint_expectation_value}
\end{equation}
which, if expanded in the same way as the direct method, directly
recovers the exact solution:
\begin{equation}
  \begin{split}
    E\{X_j\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N} b_{i_0} h_{i_0,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}
    \delta_{i_k,j} \\ &= x_{j}\:,
  \end{split}
  \label{eq:adjoint_expectation_expansion}
\end{equation}
therefore also providing an unbiased Monte Carlo estimate of the
solution.
